officedown template

### 2024-08-02 docs/verticality_future_mkd.Rmd
intro revision

# Introduuction

Vertical stacking of species from beneath the ground to the top of the canopy is an emblematic signature of biologically rich regions and a major driver of biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. In MacArthur's [-@macarthurPopulationEcologyWarblers1958] seminal work on verticality in warblers, he notes distinct partitioning of vertical niche space among species in a community. Since then, vertical stratification at local scales has been widely observed across diverse taxonomic groups [@bashamVerticalStratificationPatterns2023a]. At regional scales, patterns of community verticality vary across environmental gradients, generally increasing in wetter habitats [@bashamVerticalStratificationPatterns2023a; @scheffersIncreasingArborealityAltitude2013]. For example, verticality of frogs increases at higher elevations where cooler and wetter climates increase habitat suitability in the canopy relative to low elevations where hot and dry conditions compress frog communities toward the ground [@scheffersIncreasingArborealityAltitude2013]. Globally, verticality increases toward the warmer and wetter tropics, where vertical partitioning may facilitate high biodiversity in structurally complex forests [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019]. Despite widespread agreement that verticality influences patterns of biodiversity across spatial scales, global patterns of verticality and its environmental drivers, including climatic and vegetative factors, have not been comprehensively described for vertebrates.

## possible paragraph describing hypotheses about climate and verticality
- Vertical stratification -> higher richness because species can partition the canopy
- Therefore expect higher richness to be associated with increasing verticality
- However, climate can limits verticality
- We propose that climate limits verticality
- Where climate is suitable, canopy height will positively impact verticality
- where climate is not suitable, canopy height will have no impact on verticality

As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted. Changes in taxonomic composition of assemblages changes, are often associated with changes in functional composition. Consequently, widespread patterns of deborealization, tropicalization, and thermophilization have occurred in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. Space-for-time substitutions can be used to predict impacts of climate change on the functional composition of species assemblages. For example, the thermal tolerance of assemblages displays predictable variation across elevation gradients. As the climate warms and species' ranges shift toward higher elevations, the thermal tolerance of assemblages increases as cool-adapted species become locally extirpated and warm-adapted species immigrate into communities (a process known as thermophilization) [@feeleyClimatedrivenChangesComposition2020].

@@

Relationships between climate and verticality across spatial and temporal environmental gradients have led to predictions that verticality may decline as climates become warmer and drier (cite). While forest understories are protected from extreme climatic conditions, the canopy experiences larger fluctuations in temperature and humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. Under climatically stressful conditions, arboreal species may move to lower vertical niche positions, reducing community verticality. Temporary vertical shifts have been observed for decades across a variety of biomes [@macarthurPopulationEcologyWarblers1958; @bashamVerticalStratificationCollapses2020; @kosheleffTemperatureInfluenceActivity2009; @eppleyFactorsInfluencingTerrestriality2022]. For example, arboreal frogs in a Panamanian rainforest move toward the understory during the dry season to escape the hot and dry conditions in the canopy [@bashamVerticalStratificationCollapses2020]. Similarly, arboreal primates from the Americas to Madagascar increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022]. Yet how vertical niche shifts may alter global patterns of verticality as the climate warms is not well understood. Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe. We then examine how climate change may impact environmental conditions that have led to present day patterns of verticality.

@@

@@

# Environmental drivers of verticality

Models of mean and SES verticality indicated some similarities between the environmental drivers of mean and SES verticality across taxonomic groups, yet noticeable differences remained.
We hypothesized that canopy height would have an overall positive effect on verticality, but that its effect would vary over space, being stronger in tropical regions and lower in temperate regions, which generally have few arboreal vertebrates despite the presence of tall canopies. All of the best models included a spatially varying coefficient for canopy height. Overall, canopy height exhibited a positive effect on verticality of birds and mammals and weak positive effect or no effect on verticality of reptiles and amphibians (\@ref(fig:coef-plt). Over space the impact of canopy height was generally strongest in the Andes and Madagascar

If there is a positive relationship between canopy height and verticality, then taller canopies have more arboreal species
If there is a negative relationship between canopy height and verticality, then shorter canopies have more arboreal species
If there is not relationship between canopy height and verticality, then height of the canopy doesn't matter, potentially because climatic or evolutionary factors determine verticality.

Across taxa, precipitation of the driest and wettest months positively influenced the SES of verticality, though the impact of dry season precipitation was noticeably stronger. 
Impacts of climate on mean verticality generally mirrored those on the SES of verticality, though effects were smaller relative to the impact of canopy height.

variable effects of climate and vegetation across taxa. Mean verticality of birds and mammals was most strongly influenced by canopy height, though responses to canopy height varied across space. Though climatic variables had

Overall, the mean and SES of verticality exhibited similar responses to environmental variables across taxonomic groups. Verticality was positively associated with precipitation of the driest and wettest months and exhibited a quadratic relationship with maximum temperature of the warmest month for mean and SES verticality of birds and reptiles and SES verticality of mammals. The quadratic relationship was characterized by an initial increase in verticality followed by a decline in verticality at extremely high maximum temperatures. Mean verticality of mammals exhibited a relatively small positive relationship with maximum temperature and amphibians  except for amphibians, for which SES of verticality was negatively associated with maximum temperatures. Minimum cold temperatures positively impacted verticality of mammals, such that verticality increased as minimum temperatures increased, while having little impact on birds and mammals. Impacts of climate velocity also varied, negatively impacting birds, mammals, and amphibians, while having a small positive effect on reptiles.

*For difference section - current geographic niche underfilling and future potential to fill unused geographic space by filling currently unoccupied volume in the canopy as climate permits these habitats to become more suitable*

Though arboreal habitats remain in temperate biomes, the entire habitat volume is not occupied. This could be caused by arboreal habitats being climatically unsuitable in temperate regions and lead to a breakdown of the relationship between richness and volume. Overall, these patterns indicate a decline in the volume of habitable space toward the poles despite the structural presence of habitat.

-

    ## climate velocity represents climate instability across geographic timescales and negatively impacted verticality across vertebrate classes

-   Elevation

-   Vegetative complexity - selective logging could negatively impact verticality by reducing structural complexity

-   Decline in verticality in temperate regions despite the avialability of arboreal habitat space indicates physiological challenges of arboreal strategies in seaosnally cold and/or dry environmentals (Oliveira and Scheffers 2019)

-   for some vertebrate classes, fossoriality offers refuge from climate extremes

*plot richness by area x structural complexity and color by latitude* *then plot richness x latitude* *consider adding correlation of verticality with species richness and discuss here*

```{r}
load("./../results/amphibians_sar_vertmean2.RData")
amph.avgmod = vert.avg
amph.summ = summary(amph.avgmod)
amph.summ = round(amph.summ$coefmat.full, 2)
colnames(amph.summ) = c("Estimate", "SE", "z", "p")

load("./../results/mammals_sar_vertmean2.RData")
mammals.avgmod = vert.avg
mammals.summ = summary(mammals.avgmod)
mammals.summ = round(mammals.summ$coefmat.full, 2)
colnames(mammals.summ) = c("Estimate", "SE", "z", "p")

load("./../results/reptiles_sar_vertmean2.RData")
rept.avgmod = vert.avg
repts.summ = summary(rept.avgmod)
repts.summ = round(repts.summ$coefmat.full, 2)
colnames(repts.summ) = c("Estimate", "SE", "z", "p")

load("./../results/birdselton_sar_vertmean.RData")
birds.avgmod = vert.full
birds.summ = summary(birds.avgmod) # make sure spatialreg is loaded for summary to produce correct results
birds.summ = round(birds.summ$Coef, 2)
colnames(birds.summ) = c("Estimate", "SE", "z", "p")

@@

# Environmental drivers of verticality

## Climate

Global patterns of verticality could largely be explained by environmental drivers. To examine the impacts of climate, topography, and vegetation on global patterns of verticality, we used simultaneous autoregressive (SAR) models with a spatial error term. The best models explained explained 88-98% of the global variance in the mean and SES of mean verticality across classes (\@ref(fig:model-coefs)). However, the spatial signal in all models was strong (lambda \> 0.9), which resulted in relatively small impacts of environmental variables as well as a lower pseudo-R^2^ value when calculated based only on the predicted trends without the spatial signal.

Precipitation of the driest quarter and minimum temperatures of the coldest quarter had the largest effects on the mean and SES of verticality across vertebrate classes, both exhibiting positive effects barring the impact of minimum temperatures on amphibians, which was insignificant ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)). Maximum temperatures of the warmest quarter had opposing effects on endotherms and ectotherms, negatively impacting verticality of reptiles and amphibians while generally having positive or insignificant effects on birds and mammals, except at the highest temperatures where verticality declined ((\@ref(fig:model-coefs), \@ref(tab:mean-vert-mods), \@ref(tab:sesvert-mods)). Arboreality is therefore primarily constrained by cold minimum temperatures and limited precipitation during the dry season, while high maximum temperatures additionally limit arboreal strategies in ectotherms. The vertical climate gradient from beneath the ground to the top of the canopy drives these global constraints on arboreality. Greater climatic fluctuations in the canopy coupled with water limitation pose physiological challenges to arboreal lifestyles. For example... In contrast, terrestrial and fossorial habitats are buffered from climate extremes, which facilitates species persistence where hot, cold, or dry conditions may prevent the occurrence of arboreal lifestyles.

Dry conditions negatively impact arboreal lizard, particularly under simultaneously hot environments. <https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1301>

```         
- Though arboreal habitats can exhibit average temperatures over 2C higher than terrestrial habitats (Scheffers et al. 2013) critical thermal maxima of arboreal species may not substantially exceed that of terrestrial species (Carilo Filho et al. 2021)
```

## Historical Climate - climate velocity

Verticality is also generally higher in regions that have experienced historically stable climatic conditions, as indicated by the significant negative relationships between climate velocity and the mean verticality of birds, mammals, and amphibians and SES of verticality for mammals, as well as the absence of positive relationships for any taxa. Low climate velocity (i.e., climate stability) generally increases speciation rates and decreases extinction rates (Colville et al. 2020). In tropical forests, complex vegetation structure characterized by multilayered forests (Terborh 1985) increases opportunities for diversification among arboreal niches (see Colville et al. 2020 11-13). For example, arboreal lineages of reptiles and frogs generally exhibit higher diversification rates [@liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017], while fossorial strategies have been associated with reduced speciation and increased extinction rates in snakes [@cyriacDiggingTheirOwn2018].

-   climate stability may therefore provide the conditions necessary for diversification in regions where climatic conditions are suitable for arboreal strategies, while increased dspersal capacity of arboreal species throughout 3D arboreal environments facilitates persistence in climatically unstable regions (Sandel et al. 2011; Scheffers et al. 2017).

```{r}
bird.meanvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_meanvert.RData")
bird.meanvert = pred.df
bird.meanvert$vertvar = "mean"
bird.sesvert = load("./../results/sar_mods_forestOnly_forestSES/birds/birds_sar_sesvert.RData")
bird.sesvert = pred.df
bird.sesvert$vertvar = "ses"
bird.pred = bind_rows(bird.meanvert, bird.sesvert) %>% mutate(class = "bird")

mammals.meanvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_meanvert.RData")
mammals.meanvert = pred.df
mammals.meanvert$vertvar = "mean"
mammals.sesvert = load("./../results/sar_mods_forestOnly_forestSES/mammals/mammals_sar_sesvert.RData")
mammals.sesvert = pred.df
mammals.sesvert$vertvar = "ses"
mammals.pred = bind_rows(mammals.meanvert, mammals.sesvert) %>% mutate(class = "mammals")

rept.meanvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_meanvert.RData")
rept.meanvert = pred.df
rept.meanvert$vertvar = "mean"
rept.sesvert = load("./../results/sar_mods_forestOnly_forestSES/reptiles/reptiles_sar_sesvert.RData")
rept.sesvert = pred.df
rept.sesvert$vertvar = "ses"
rept.pred = bind_rows(rept.meanvert, rept.sesvert) %>% mutate(class = "rept")

amph.meanvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_meanvert.RData")
amph.meanvert = pred.df
amph.meanvert$vertvar = "mean"
amph.sesvert = load("./../results/sar_mods_forestOnly_forestSES/amphibians/amphibians_sar_sesvert.RData")
amph.sesvert = pred.df
amph.sesvert$vertvar = "ses"
amph.pred = bind_rows(amph.meanvert, amph.sesvert) %>% mutate(class = "amph")

pred = bind_rows(bird.pred, mammals.pred, rept.pred, amph.pred)

# Impacts of climate change on verticality

To examine the potential impacts of climate change on the verticality of vertebrate assemblages, we used the coefficients from the SAR models to predict assemblage verticality for the time period 2071-2100 using the SSP585 future climate scenario, which represents the upper boundary of predicted warming under a scenario of fossil-fueled development. Though climate change will impact vegetation structure, we were unable to incorporate predictions of future vegetation into the models due to a lack of global future predictions of vegetation height and density. Because we could not predict the spatial signal in verticality, we compared predictions under future climate conditions to predictions under present climate conditions that do not incorporate the spatial signal. - We find that x% of communities could experience declines in verticality. - Declines in verticality are concentrated in the tropics. - Between 23 and -23 latitude, x% birds, x% mammals, x% reptiles, and x% amphibians exhibit declines in verticality - Declines in the Amazon were notably severe

@@

-   What causes changes in verticality - vertical niche shift, immigration, or local extinction?

@@

-   In temperate regions, we find...

-   Predicted changes in verticality led to homogenization of vertical strategies globally.

@@

In the tropics, the use of arboreal habitat likely contributes to high species richness (cite Oliveira, also see Naikatini et al. 2022, MacArthur). The large volume of habitat in tropical biomes facilitates species packing within a given area as well as expansion across vertically stratified resources. However, habitat volume and structural diversity is generally not related to species richness in forested ecosystems in the US, except for birds (La Rue et al. 2023). Low arboreality toward higher latitudes may explain the breakdown in the volume-richness relationship and be caused by low climatic suitability of arboreal habitats in temperate climates, except for birds which continue to use arboreal habitats at northern latitudes.

-increasing arboreality at norther latitudes suggests that climate change could lead to arboreal habitats in boreal regions becoming more suitable for species, thereby releasing competitive pressure as species shift poleward by providing more habitat volume. -niche attrition - land use change eliminates the volume of arboreal habitats, which leads to declines in arboreality. Will climate change cause further attrition of arboreal habitats?? - elevator to extinction for mountaintop species = digging a grave

-   vertical niche shifts occur in response to novel habitat, global change, etc.
-   e.g., shift in birds to arboreal foraging on islands (Lapiedra et al. 2021 Niche shifts after island colonization...)

Despite the presence of biomes with vertical structure in temperate forests the proportion of arboreality steeply declines, suggesting that the geographic niche space is underfilled by the animal community.

Low climatic suitability may limit arboreality and thus diversity in temperate biomes.

## The influence of vegetation
Indeed, it is expected that the volume of habitat space explain variation in species richness among forested biomes. However, this relationship is not upheld for birds and mammals (LaRue et al. 2023). Declining arboreality in temperate forests that have the vegetation structure to support arboreal strategies may explain this pattern and be caused by environmental factors limiting the suitability of canopy habitats.

>>>>

# Introduction

Vertical stacking of species is an emblematic signature of biologically rich regions and influences biogeographic patterns from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. These climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. Stratification across this vertical climate gradient may reduce competition and increase species richness (*cite*). Tall, multi-layered forests are thus expected to support both highly diverse and highly arboreal species assemblages by increasing the volume of habitat and enabling species to finely partition the vertical forest structure [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. These conditions occur in tropical forests where high proportions of arboreal species may contribute to their biologically rich assemblages [@macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019]. However, vegetation structure often fails to predict patterns of species richness at continental to global scales [@rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023].

At global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests. For example, low minimum temperatures in the canopies of high latitude forests may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. Similarly, exposure to hot and dry conditions in the canopies of tropical lowland forests may reduce the suitability of arboreal habitats for heat or desiccation sensitive species, such as frogs [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020]. Release of these climatic constraints at higher elevations, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013]. We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological constraints.

Similarly, spatial variation may occur in the relationship between species richness and assemblage verticality, defined here as the average vertical position of species in an assemblage [@oliveiraVerticalStratificationInfluences2019]. While verticality and richness are often assumed to exhibit a positive relationship, the strength of this relationship is variable [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*. In regions where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weka because species are unable to vertically partition habitat. In contrast, when an increase in canopy height allows the addition of arboreal species to the assemblage, increasing assemblage verticality, we expect a strong positive relationship between verticality and species richness.

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and result in either shifts toward greater use of lower vertical positions or local extirpation [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. Both scenarios would reduce assemblage verticality, reflecting climate-induced shifts in assemblage thermal tolerance caused by local extirpation or range shifts [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality. However, the extent to which climate change may alter global patterns of verticality is unknown.

Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates. To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes. We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness. Finally, we predict how climate change may impact global patterns of verticality.

Long-term climate instability may additionally impact global patterns of verticality [@scheffersVerticalArborealityHorizontal2017]. Physiological adaptations to climate variability in the canopy may confer tolerance to extreme hot and dry conditions [@tracyNotJustSmall2010; @leahyArborealityDrivesHeat2022]. Alternatively, vertical dispersal between the ground and canopy increases opportunities for behavioral adaptations to climate fluctuations [@scheffersIncreasingArborealityAltitude2013; @scheffersVerticalArborealityHorizontal2017; @bashamVerticalStratificationCollapses2020]. However, observed impacts of historical climate instability on assemblage verticality are inconsistent [@scheffersVerticalArborealityHorizontal2017; @oliveiraVerticalStratificationInfluences2019].

As climate change causes the reorganization of species assemblages globally [@scheffersBroadFootprintClimate2016], patterns of verticality may be impacted. Functional changes to community composition are often detected in response to climate change. For example, deborealization, tropicalization, and thermophilization have occurred as community thermal tolerances increase in response to warming temperatures [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. Warming and drying climates may cause arboreal species to seek refuge at lower positions within the vertical forest structure, disperse to cooler climates toward higher latitudes or elevations, or become locally extinct. These processes will alter assemblage verticality. For example, verticality may decline in tropical lowlands where species are increasingly exposed to extreme hot and dry conditions [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. In contrast, verticality may increase toward higher latitudes or elevations where warming temperatures may release constraints on arboreality. While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.

@@

Despite widespread agreement that verticality influences patterns of biodiversity and vulnerability to climate change, global patterns of verticality and its environmental drivers have not been comprehensively described for vertebrates. Here, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with global climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes across the globe. We then examine how climate change may impact assemblage verticality.

@@

To validate the model, we examined quantile residuals to ensure normality and lack of spatial pattern.

@@

## Environmental drivers of verticality

We expected verticality to increase with increasing canopy height, vegetation density, precipitation, and minimum cold temperatures. While the models were generally consistent with these hypotheses, there were some notable differences.

### need a section on model fit ###
Models predicted verticality well. However, much of the predictive ability was due to the spatial random field. When the spatial random field was excluded from model predictions, models generally underestimated verticality in regions with high verticality and overestimated verticality in regions with low verticality.

### Climate

The models generally indicated that verticality was limited by low precipitation, high maximum temperatures, and low minimum temperatures, though some variation occurred between taxa. Additionally, climatic impacts relative to the impact of canopy height was stronger for SES verticality than mean verticality.

Across taxa, precipitation of the driest and wettest months positively influences mean and SES verticality, though the impact of dry season precipitation on SES verticality was noticeable stronger (\@ref(fig:mod-coefs)). Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality. Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile verticality, as well as SES verticality of mammals. This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further. Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship. Climate velocity was not included in the best models for mammals verticality or for SES verticality of amphibians. For the remaining models, climate velocity (i.e., climate instability) negatively impacted bird verticality and mean verticality of amphibians and positively impacted reptile verticality.

### Vegetation

We hypothesized that canopy height and vegetation density would positively impact verticality, but that the impact of canopy height would vary across space. Specifically, we expected canopy height to have a strong positive effect on verticality in warmer and wetter regions that support arboreal assemblages and a negligible effect in regions with climate conditions that support fewer arboreal species.

Globally, canopy height positively impacted verticality for mammals and birds and had little impact on verticality of reptiles and amphibians, while vegetation density had a negligible impact on verticality of all taxa (\@ref(fig:coef-plt). The best models for all taxa included a spatially varying coefficient for canopy height, which caused the wide confidence interval in the global estimate its coefficient.

Spatial variation in the effect of canopy height was greater for models of SES verticality than for mean verticality, though both exhibited similar patterns (\@ref(fig:svc-map)). For birds, canopy height exhibited the strongest positive effect on SES verticality in the Andes, central America, Indomalayan moist deciduous forests, the forest-savanna woodlands and thickets of eastern Africa, and northern taiga forests (\@ref(fig:svc-map)A,E). Spatial variability of the coefficient of canopy height for models of SES verticality exhibited similar pattern in mammals. However, strong positive effects additionally occurred in the south-eastern US, subpolar forests of South America, Guinean forest-savanna, and Australian temperate savanna and woodlands (\@ref(fig:svc-map)B,F). For reptiles, similar regions showed weaker positive relationships between canopy height and SES verticality. However, Madagascar exhibited a strong positive coefficient (\@ref(fig:svc-map)C,G). The coefficient of canopy height was also strongly positive for Amphibians in Madagascar as well as Sri Lanka relative to other regions (\@ref(fig:svc-map)D,H). Notably, the coefficient of canopy height was near zero in tropical broadleaf forests across the globe and was not negligible in cooler regions, contrary to our predictions.

## Climate change impacts

```{r}
# calculate percent change in verticality
calc_change = function(pred.f, taxa, response) {
  pred.f = pred.f %>% 
    dplyr::select(x,y,est.dif) %>% 
    mutate(taxa = taxa,
         response = response)
return(pred.f)
}

load("./../results/sdmTMB_models/amphibians_meanvert.RData")
amph.meanvert = calc_change(pred.f, taxa = "Amphibians", response = "Mean Verticality")

load("./../results/sdmTMB_models/amphibians_sesvert.RData")
amph.sesvert = calc_change(pred.f, taxa = "Amphibians", response = "SES Verticality")

load("./../results/sdmTMB_models/reptiles_meanvert.RData")
rept.meanvert = calc_change(pred.f, taxa = "Reptiles", response = "Mean Verticality")

load("./../results/sdmTMB_models/reptiles_sesvert.RData")
rept.sesvert = calc_change(pred.f, taxa = "Reptiles", response = "SES Verticality")

load("./../results/sdmTMB_models/mammals_meanvert.RData")
mammals.meanvert = calc_change(pred.f, taxa = "Mammals", response = "Mean Verticality")

load("./../results/sdmTMB_models/mammals_sesvert.RData")
mammals.sesvert = calc_change(pred.f, taxa = "Mammals", response = "SES Verticality")

load("./../results/sdmTMB_models/birds_meanvert.RData")
birds.meanvert = calc_change(pred.f, taxa = "Birds", response = "Mean Verticality")

load("./../results/sdmTMB_models/birds_sesvert.RData")
birds.sesvert = calc_change(pred.f, taxa = "Birds", response = "SES Verticality")

dat = rbind(amph.meanvert, amph.sesvert, rept.meanvert, rept.sesvert, mammals.meanvert, mammals.sesvert, birds.meanvert, birds.sesvert)

dat.change = dat %>% 
  group_by(taxa, response) %>% 
  summarise(max = max(est.dif), min = min(est.dif))

dat.change.ses = dat.change %>% filter(response == "SES Verticality")
dat.change.mean = dat.change %>% filter(response == "Mean Verticality")

ses.poschange = dat.change.ses[which(dat.change.ses$max == max(dat.change.ses$max)),]
ses.negchange = dat.change.ses[which(dat.change.ses$min == min(dat.change.ses$min)),]

mean.poschange = dat.change.mean[which(dat.change.mean$max == max(dat.change.mean$max)),]
mean.negchange = dat.change.mean[which(dat.change.mean$min == min(dat.change.mean$min)),]

@@

Changes in verticality in response to climate change were small relative to the total variation in mean and SES verticality. The largest changes in SES verticality were observed for birds, which exhibited a maximum decline of `r round(as.numeric(ses.negchange[1, "min"]),2)` and a maximum increase of `r round(as.numeric(ses.poschange[1, "max"]),2)`. The larges changes in mean verticality were observed for amphibians, which exhibited a maximum decline of `r round(as.numeric(mean.negchange[1, "min"]), 2)` and a maximum increase of `r round(as.numeric(mean.poschange[1, "max"]), 2)`. For birds and mammals, mean and SES verticality generally declined in tropical moist broadleaf forests globally in addition to the temperate forests, savannas, and woodlands of southern Australia and the Brazilian Atlantic forests. In contrast, verticality primarily increased in the boreal forests of Canada and northern Eurasia, tropical and subtropical savannas and woodlands in Africa, and tropical savannas in northern Australia (\@ref(fig:vert-dif-biome)). Changes in verticality of Reptiles and Amphibians were characterized by declines in most regions. However, amphibian verticality was expected to experience minor increases in African tropical and subtropical savannas and woodlands (\@ref(fig:vert-dif-biome)).

@@

@@

@@
